---
title: "VWC percentielen"
author: "Iris Eekhout"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(gseddata)
library(dplyr)
library(tidyr)
library(ggplot2)

NLD <- data.frame(gseddata::get_data(cohorts = c(53,54)))

theme_set(theme_light())
```

## Approaches

There are three possible approaches:

(1) Fully empirically by just computing the percentage for each age in the data then approximate the percentile ages from that table.
(2) Use a logistic regression to estimate the association between age and probability to pass for each item. Use the logistic function to estimate the ages at the percentiles
(3) Use a GAMLSS model to estimate the association between age and probability to pass.
(4) Use the D-score model and references


### Empirical percentages

```{r}


empirical <-
NLD %>%
  mutate(ageweeks = (agedays + 1) / 7,
         agemos = agedays/ 365.25 *12) %>%
  mutate(week = ceiling(ageweeks),
         month = ceiling(agemos)) %>%
  pivot_longer(cols = starts_with("ddi"), names_to = "item", values_to = "score") %>%
  group_by(item, week) %>%
  summarise(n = sum(!is.na(score)),
            p = sum(score, na.rm = T)/n,
            .groups = "drop") %>%
  left_join(dscoreddi::itemtableVWO %>% select(ID.VWO2005, labelNL, item), by = "item") %>%
  select(item, ID.VWO2005, labelNL, week, n, p)

percentile_age <- 
empirical %>%
  group_by(item) %>%
  summarise(
    A10 = approx(p, week, xout = 0.10)$y / 52 * 12,
    A50 = approx(p, week, xout = 0.50)$y / 52 * 12,
    A90 = approx(p, week, xout = 0.90)$y / 52 * 12)


pref_emp <- dscoreddi::prefdat %>%
  select(-c(A10, A50, A90)) %>%
  left_join(percentile_age, by = "item")



emp <- empirical %>% filter(n > 10) %>% mutate(agedays = round(week * 7),0) %>% select(item, agedays, p)

```


### Logistic regression


```{r}
items <- colnames(NLD)[grepl("^ddi", colnames(NLD))]
df <- data.frame(agedays = 1:1461)
logistic_age <- data.frame()
for(i in items){
  
  form <- as.formula(paste(i, "~ agedays"))
  fit <- glm(form, data = NLD, family = "binomial")
  df$prob <- predict.glm(fit, df, type = "response")
  
  out <- 
  data.frame(item = i, 
             A10 = approx(df$prob, df$agedays, xout = 0.1)$y /365.25 *12,
             A50 = approx(df$prob, df$agedays, xout = 0.5)$y /365.25 *12,
             A90 = approx(df$prob, df$agedays, xout = 0.9)$y /365.25 *12
  ) 
  logistic_age <- bind_rows(logistic_age, out)
}

pref_log <- dscoreddi::prefdat %>%
  select(-c(A10, A50, A90)) %>%
  left_join(logistic_age, by = "item")

```

### D-score model


```{r}
      
   scalefactor <- 2.09
   p <- c(1:100)
   reference <- dscoreddi::expanded_reference
   
   ib <- dscoreddi::itembank_vwc
   pd <- matrix(ib$tau, nrow = nrow(ib), ncol = length(p)) + 
        matrix(scalefactor * qlogis(p/100), nrow = nrow(ib), 
            ncol = length(p), byrow = TRUE)
   pa <- approx(x = reference$mu, y = reference$age, xout = as.vector(pd))$y*365.25 
   pa <- matrix(pa, ncol = length(p))
   pa <- data.frame(ib[, "item"], round(pa,0))
   names(pa) <- c("item", paste0("A", p))
   
   dipa <- pa %>% pivot_longer(-1, names_to = "p", values_to = "agedays") %>% mutate(p = as.numeric(gsub("A","", p))/100)
   
   
   
   #ggplot(ipa, aes(agedays, p))+
   #   geom_line()+
   #   facet_wrap(~item)
    
    
```



### GAMLSS model

```{r eval=FALSE, include=TRUE}
library(gamlss)
library(dscoreddi)

dfout <- data.frame()


items <- colnames(NLD)[grepl("^ddi", colnames(NLD))]
keep <- sapply(NLD[, items], function(x) min(table(x)) >= 10)
items <- items[keep]

df <- data.frame(agedays = 7:1461)
for(i in items){
# i <- "ddicmm033" 
    df2 <- df %>% mutate(item = i)
     
    x <- NLD %>% select(agedays, !!i) %>% na.omit
    
    form <- as.formula(paste(i, "~ agedays"))
    fit <- glm(form, data = x, family = "binomial")
    df2$logistic <- predict.glm(fit, newdata = df, type = "response")
     
    formd <- as.formula(paste(i, "~ ps(agedays)"))
		gam.default <- gamlss(formd, data=x, family = BI, na.action = na.omit)
    df2$gam_dfd <- predict(gam.default, "mu", newdata = df, type = "response")
		
    form2 <- as.formula(paste(i, "~ ps(agedays, df = 2)"))
		gam.df2     <- gamlss(form2, data=x, family = BI)
		df2$gam_df2 <- predict(gam.df2, "mu", newdata = df, type = "response")
		
	  form2.5 <- as.formula(paste(i, "~ ps(agedays, df = 2.5)"))
    gam.df2.5   <- gamlss(form2.5, data=x, family = BI, na.action = na.omit)
	  df2$gam_df2.5 <- predict(gam.df2.5, "mu", newdata = df, type = "response")
		
	  form3 <- as.formula(paste(i, "~ ps(agedays, df = 3)"))
	  gam.df3     <- gamlss(form3, data=x, family = BI, na.action = na.omit)
	  df2$gam_df3 <- predict(gam.df3, "mu", newdata = df, type = "response")
		
	 
    df3 <- df2 %>%
      pivot_longer(-c(agedays, item), names_to = "model", values_to = "p")
    
    ggplot(data = df3)+
      geom_line(aes(x = agedays, y = p, color = model, group = model), size = 2, alpha = 0.5 )+
      ylim(c(0,1))+
      geom_abline(aes(intercept = 0.1, slope = 0))+
      geom_abline(aes(intercept = 0.5, slope = 0))+
      geom_abline(aes(intercept = 0.9, slope = 0)) +
      facet_grid(model~.)
    
    dfout <- bind_rows(dfout, df3)
      
    
}
 

     
    dsc <- dipa %>% mutate(model = "p_dscore")
    empp <- emp %>% mutate(model = "empirical")
                    
   # dfout <- dfout %>% bind_rows(dsc) #%>%
      #bind_rows(empp)
           
 ggplot(data = dfout)+
      geom_line(aes(x = agedays, y = p, color = model, group = model), size = 1.2, alpha = 0.5 )+
      ylim(c(0,1))+
      geom_abline(aes(intercept = 0.1, slope = 0))+
      geom_abline(aes(intercept = 0.5, slope = 0))+
      geom_abline(aes(intercept = 0.9, slope = 0)) +
      facet_wrap(~item)
    


empest <- percentile_age %>% pivot_longer(A10:A90, names_to = "perc", values_to = "agemos") %>%
  mutate(p = as.numeric(gsub("A", "", perc))/ 100,
         agedays = round(agemos / 12 * 365.25,0))

plots <- list()

for(i in items){
  lab <- dscoreddi::itemtableVWO %>% filter(item == i) %>% pull(labelNL)
  
  p <- 
   ggplot(data = dfout %>% filter(item == i))+
      geom_line(aes(x = agedays, y = p, color = model, group = model), size = 1.2, alpha = 0.5 )+
      geom_point(data = empp %>% filter(item == i), aes(x = agedays, y = p), color= "darkgrey" )+
      ylim(c(0,1))+
      geom_point(data = empest %>% filter(item == i), aes(x = agedays, y = p), size = 2)+
      geom_abline(aes(intercept = 0.1, slope = 0))+
      geom_abline(aes(intercept = 0.5, slope = 0))+
      geom_abline(aes(intercept = 0.9, slope = 0)) +
    ggtitle(paste("item:", i, "-", lab))
 plots[[i]] <- p 
}

#Export
pdf('results/vwc_approaches_pt_geend.pdf',height = 6, width = 8)
plots
dev.off()
    
```



```{r}
pref <- bind_rows(
  pref_emp %>% mutate(computation = "empirical"), 
  pref_log %>% mutate(computation = "logistic"), 
  dscoreddi::prefdat %>% mutate(computation = "dscore")
)


  
```


```{r, fig.height=11, fig.width=10}

 ggplot(data = pref %>%
                     filter(sideA == 1), #selecteer milestons met A10 < 15 maanden
        aes(x = reorder(labelNLn, -nr),  y= A50, color = computation, group = computation)
                   ) + #changed
            geom_point()+
            geom_errorbar(aes(ymin = A10, ymax = A90), size = 1, width = 0.2, alpha = 0.6)+
            
            coord_flip()+ xlab("") +
            theme(legend.position = "top") +
  
          scale_y_continuous(name = "", breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15, seq(15,51,3)), minor_breaks = c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,16,17,19,20,22,23,25,26,28,29,31,32,34,35,37,38,40,41,43,44,46,47,49,50), limits= c(0,18),position = "right", sec.axis= dup_axis(name = "Leeftijd (maanden)"), expand = c(0,0.1))
  
  
```


```{r, fig.height=11, fig.width=10}

  

 ggplot(pref %>%
                     filter(sideB == 1),  #selecteer milestons met A10 < 15 maanden
                   aes(x = reorder(labelNLn, -nr),  y= A50, color = computation, group = computation)) + #changed
            geom_point()+
            geom_errorbar(aes(ymin = A10, ymax = A90), size = 1, width = 0.2, alpha =0.5)+
            
            coord_flip()+ xlab("") +
            theme(legend.position = "top") +

          scale_y_continuous(name = "", breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15, seq(15,51,3)), minor_breaks = c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,16,17,19,20,22,23,25,26,28,29,31,32,34,35,37,38,40,41,43,44,46,47,49,50), limits= c(8,50),position = "right", sec.axis= dup_axis(name = "Leeftijd (maanden)"), expand = c(0,0.1))
  
  
```


Het lijkt erop dat de empirische data zorgt voor erg brede intervallen voor sommige items. 

Ook vergelijken met Denver data?



## Per domein

Fijne motoriek

```{r, fig.height=11, fig.width=10}
   ggplot(pref %>% filter(domein == "Fijne motoriek"), #selecteer milestons met A10 < 15 maanden
               aes(x = reorder(labelNLn, -nr),  y= A50, group = computation, color = computation, label = labelNL)) +
        geom_point()+
        geom_errorbar(aes(ymin = A10, ymax = A90), size = 1, width = 0.2, alpha =0.5)+
        coord_flip()+
     theme_light() +
     theme(legend.position = "top") +
        scale_y_continuous(name = "", breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15, seq(15,51,3)), minor_breaks = c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,16,17,19,20,22,23,25,26,28,29,31,32,34,35,37,38,40,41,43,44,46,47,49,50), limits= c(0,50),position = "right", sec.axis= dup_axis(name = "Leeftijd (maanden)"), expand = c(0,0.1))+
        xlab("")+
        ggtitle("Fijne motoriek/Adaptie/Persoonlijkheid en Sociaal gedrag")

```


Grove motoriek

```{r, fig.height=11, fig.width=10}
   ggplot(pref %>% filter(domein == "Grove motoriek"), #selecteer milestons met A10 < 15 maanden
               aes(x = reorder(labelNLn, -nr),  y= A50, group = computation, color = computation, label = labelNL)) +
        geom_point()+
        geom_errorbar(aes(ymin = A10, ymax = A90), size = 1, width = 0.2, alpha =0.5)+
        coord_flip()+
        theme(legend.position = "top") +
        scale_y_continuous(name = "", breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15, seq(15,51,3)), minor_breaks = c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,16,17,19,20,22,23,25,26,28,29,31,32,34,35,37,38,40,41,43,44,46,47,49,50), limits= c(0,50),position = "right", sec.axis= dup_axis(name = "Leeftijd (maanden)"), expand = c(0,0.1))+
        xlab("")+
        ggtitle("Grove motoriek")

```


Communicatie

```{r, fig.height=11, fig.width=10}
   ggplot(pref %>% filter(domein == "Communicatie"), #selecteer milestons met A10 < 15 maanden
               aes(x = reorder(labelNLn, -nr),  y= A50, group = computation, color = computation, label = labelNL)) +
        geom_point()+
        geom_errorbar(aes(ymin = A10, ymax = A90), size = 1, width = 0.2, alpha =0.5)+
        coord_flip()+
        theme(legend.position = "top") +
        scale_y_continuous(name = "", breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15, seq(15,51,3)), minor_breaks = c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5,10.5,11.5,12.5,13.5,14.5,16,17,19,20,22,23,25,26,28,29,31,32,34,35,37,38,40,41,43,44,46,47,49,50), limits= c(0,50),position = "right", sec.axis= dup_axis(name = "Leeftijd (maanden)"), expand = c(0,0.1))+
        xlab("")+
        ggtitle("Communicatie")

```

